---
title: "US Covid Rates"
author: "Donald Jackson"
date: '`r format.Date(Sys.Date(), "%B %d %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
params:
  highlightState:
    input: text
    label: State to highlight
    value: Massachusetts
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
											message = FALSE,
											fig.path = paste(here::here(), "results", "us", "", sep = "/"))

# get state to highlight
highlightState <-  params$highlightState
```

# Objective

Generate exploratory visualizations for US COVID-19 infection and death rates using data from the New York Times.

# Procedure

1. Load current state-level case and death data
1. Load current county-level case and death data
1. Load state and county-level population data from 2019 census and join to case and death datasets
1. Load map coordinates for chloropleth maps
1. Plot cases per capita at state level
1. Map cases per capita at county level
1. Plot deaths per capita at state level
1. Plot deats versus cases at state level

```{r packages}

suppressPackageStartupMessages({
	library(covidPlots)
	library(envDocument)
	library(here)
	library(ggToys)
	library(dplyr)
	library(drake)

})

```

```{r set_defaults}
# set default ggplot theme
ggplot2::theme_set(theme_dj())

```

## Data


```{r load_caseData}
# load state and county case data from drake cache

drake::loadd(stateData)

drake::loadd(countyData)

```

Using state-level case data downloaded from
`r attr(stateData, "source")` at
`r format.POSIXct(attr(stateData, "timestamp"), "%B %d %Y %T")`
and county-level case data downloaded from
`r attr(countyData, "source")` at
`r format.POSIXct(attr(countyData, "timestamp", "%B %d %Y %T"))`.



```{r subset_last_Data}
lastStateData <- getLastStateData(stateData)

lastCountyData <- getLastCountyData(countyData)

```

```{r load_map}
# load us map as json for chloropleth map
# usMap <- getUsMap()

loadd(usMap)

```

# Results

## State-Level

### Cases per Capita
```{r cases_byStateDate}

p_casesByState <- plotCurrentCount(lastStateData, xCol = "state",
																	 value = "cases_per_capita",
																	 hGroupName = "byState")

p_caseGrowthByState <- plotEventVTime(stateData, value = "cases_per_capita",
																			hGroupName = "byState")

plotly::subplot(p_casesByState, p_caseGrowthByState,
								widths = c(0.25, 0.75),
								shareX = F, shareY = T,
								which_layout = 1) %>%
	plotly::highlight(dynamic = T, selectize = T, persistent = F,
										on = "plotly_click",
										defaultValues = highlightState)

```

```{r new_cases_by_state}
# plot new cases as rolling average

p_newCasesState <- plotNewEvents(stateData, event = "new_cases", 
																 hGroupName = "byState") %>%
	plotly::highlight(dynamic = T, selectize = T, persistent = F,
						on = "plotly_click",
						defaultValues = highlightState)

p_newCasesState

```

```{r case_doubling, eval = T}
# Plot case doubling times
dtPlot <- plotDoublingTime(stateData)


# growthRate

dtPlot <- dtPlot %>%
	plotly::highlight(dynamic = T, selectize = T, persistent = F,
						on = "plotly_click",
						defaultValues = highlightState)

dtPlot
```

### US Case Map

```{r caseMap, eval = T}

p_caseMap <- plotUsMap(lastCountyData, event = "cases_per_capita", map = usMap)
p_caseMap 

```

### Deaths per Capita
```{r deaths_byStateDate}
p_deathsByState <- plotCurrentCount(lastStateData, xCol = "state",
																	 value = "deaths_per_capita",
																	 hGroupName = "byState")

p_deathGrowthByState <- plotEventVTime(stateData, value = "deaths_per_capita",
																			hGroupName = "byState")

plotly::subplot(p_deathsByState, p_deathGrowthByState,
								widths = c(0.25, 0.75),
								shareX = F, shareY = T,
								which_layout = 1) %>%
	plotly::highlight(dynamic = T, selectize = T, persistent = F,
										on = "plotly_click",
										defaultValues = highlightState)

```

```{r new_deaths_by_state}
# plot new deaths as rolling average

p_newDeathsState <- plotNewEvents(stateData, event = "new_deaths", 
																 hGroupName = "byState") %>%
	plotly::highlight(dynamic = T, selectize = T, persistent = F,
						on = "plotly_click",
						defaultValues = highlightState)

p_newDeathsState

```

<!-- abstract in second yaml block - needs new line before and after separator plus indent -->

```{r abstract_values, include = FALSE}
# get summary values for abstract

todaysData <- getLastStateData(stateData)

# total new cases and deaths for US
us_values <- todaysData %>% summarize(
	new_cases = sum(new_cases),
	new_deaths = sum(new_deaths)
)

# total new cases and deaths for selected state
hStateData <- filter(todaysData, state == highlightState)

```

---
abstract: |
	On `r format.Date(Sys.Date(), "%B %d %Y")` the US reported 
	`r us_values$new_cases` new COVID-19 cases
	and `r us_values$new_deaths` new deaths.  `r highlightState` reported
	`r hStateData$new_cases` new cases and `r hStateData$new_deaths` deaths.
---

# System information
```{r environment, results = "asis"}
# summarize environment and session info
env_doc("table")
```